\revision{Confirmation of \srcfull{} as a long period radio transient}{A new long period radio transient: Discovery of pulses repeating every 1.16 hours from \srcfull{}}

### 2025-07-10 paper/main.tex
Completed draft revision after getting first referee report back

\newcommand{\todo}[1]{{\color{red}#1}}
\newcommand{\src}{J1755$-$2527}
\newcommand{\srcfull}{ASKAP J175534.9$-$252749.1}

@@

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
\title[Confirmation of \src{} as LPT]{Confirmation of \srcfull{} as a long period radio transient}

@@

% Abstract of the paper
\begin{abstract}
We report the discovery of several new pulses from the source \srcfull{} (\src{}), originally identified from a single 2-min long pulse, confirming it as a long period transient (LPT) with a period of ${\sim}1.16\,$hours. The pulses are significantly scattered, consistent with Galactic electron density models. Two of the new pulses also had measurable polarisation, but unlike the originally detected pulse, the polarisation angle does not behave as expected from the rotating vector model. We interpret historical non-detections of \src{} as an intrinsic intermittency that occurs on month-long timescales, and discuss possible causes. We conjecture that, like some other LPTs with periods $\gtrsim 1\,$hour, \src{} is a white-dwarf binary pulsar, but note that its period is marginally shorter than the minimum known orbital period for the class of cataclysmic variables known as polars.
\end{abstract}

@@

\srcfull{} (hereafter, \src{}) is a transient radio source discovered \citep[][hereafter \citetalias{2024MNRAS.535..909D}]{2024MNRAS.535..909D} in the Variables and Slow Transients (VAST) survey \citep{2013PASA...30....6M}.
A single, highly polarised pulse was observed, lasting approximately 2 minutes, exhibiting a steep spectrum ($\alpha = -3.1$, for flux density $S_\nu\propto\nu^\alpha$).
Extensive follow up in new and archival data at multiple wavelengths turned up no other pulses.
Despite this, the authors concluded that \src{} was most likely a long period transient (LPT), a newly emerging class of radio source with periods ranging from minutes to hours \citep[see e.g.][]{2022Natur.601..526H,2023Natur.619..487H,2024NatAs...8.1159C}.

Although optical observations have provided compelling evidence that at least some LPTs are detached, possibly strongly magnetic, white dwarf (WD) binaries \citep{deRuiter2025,2024ApJ...976L..21H,2025A&A...695L...8R}, it remains unclear if this model can account for all known LPTs \citep[e.g.][]{2022ApJ...940...72R,Lee2025}.
Optical follow-up is challenging for \src{}, which is situated very close to the Galactic plane ($b = -0\overset{\circ}{.}12$), at a distance of ${\sim}4.7\,$kpc \citepalias{2024MNRAS.535..909D}.
Unsurprisingly, archival searches at \src{}'s location did not turn up any source down to ${\sim}23$ AB mag.
Without more detections in the radio or at other wavelengths, it seemed unlikely that the nature of \src{} could ever be resolved.

@@

Pulses from \src{} were detected with the Murchison Widefield Array \citep[MWA;][]{Tingay2013}, the Australian SKA Pathfinder \citep[ASKAP;][]{2021PASA...38....9H}, MeerKAT \citep{2016mks..confE...1J}, and the Australia Telescope Compact Array \citep[ATCA;][]{cabb}, all radio interferometers capable of imaging.
A summary of all the observations containing pulse detections included in this work is summarised in \Tab~\ref{tbl:obs}.
In this section, we describe how each telescope was used to generate dynamic spectra of \src{}.

@@

\src{} was detected blindly with the Murchison Widefield Array \citep[MWA;][]{Tingay2013} in observations taken between June--September 2024 under the `Galactic Plane Monitor', project code G0080. The prime objective of the campaign was to search for long-period transients by scanning the Galactic plane over $285^\circ < l < 65^\circ$, $|b| < 15^\circ$ on a bi-weekly cadence at 185--215\,MHz \citep[see Methods of][]{2023Natur.619..487H}; a full description will be released by Hurley-Walker et al., in prep. We also observed over June--September 2022 but made no detections in these data, as reported in \citetalias{2024MNRAS.535..909D}.
Subsequent to its redetection, we commenced a Director's Discretionary Time (DDT) campaign under project code D0042, to obtain further observations, adjusting the frequency from 200.32\,MHz to 184.96\,MHz for some of the observations to attempt to improve measurements of its DM.
Time of arrival (ToA) predictions for the DDT observations were made using a preliminary ephemeris derived from the GPM detections, which, by virtue at being at the same or similar frequencies, absorbed the delays due to dispersion and scattering into the `PEPOCH' term.

@@

With the re-detection of the source in the MWA data, and the calculation of an ephemeris (\Sect~\ref{sec:analysis}), we re-reduced the data in which we would predict that \src{} would appear. We made one further detection, in SBID~49153, at a relatively low significance ($\sim5\sigma$) compared to the blind detections. We followed a similar procedure as the above to generate dynamic spectra for each of the observations, using a minimum baseline length of 500\,m to suppress uncleaned Galactic diffuse emission.

@@

The correlator data were calibrated using the standard SARAO SDP calibration pipeline, and imaged using \textsc{WSClean}. We formed a deep model of the sources within the primary beam, excepting \src{}, and subtracted this from the visibilities. We then averaged the visibilities, excluding all baselines shorter than 150\,m, to produce a dynamic spectrum of the source, shown in \Fig~\ref{fig:stacked_spectra}.

@@

These ToAs were then barycentred and fit for both period and DM, resulting in the values given in \Tab~\ref{tbl:ephemeris}.
The residuals are presented in \Fig~\ref{fig:pulse_details} and \Tab~\ref{tbl:toas}, along with other fitted properties of the pulses.
We also tried fitting a timing model that included a spin-down parameter, and found that it was consistent with zero spin-down to within $1\sigma$, and so is not included in the ephemeris.

\begin{table}
  \centering
  \caption{Timing ephemeris for \src{}}
  \label{tbl:ephemeris}
  \begin{tabular}{lc}
    \hline
    Parameter & Value \\
    \hline
    Period (s) & $4186.3285 \pm 0.0002$ \\
    PEPOCH (MJD) & $59965.03792 \pm 0.00003$ \\
    DM (pc cm$^{-3}$) & $733 \pm 22$ \\
    $\tau_{\rm sc,1 GHz}$ (s) & ${\sim}0.07$ \\
    \hline
  \end{tabular}
\end{table}

@@

\src{}'s period and duty cycle are not dissimilar to ILT J1101+5521 and GLEAM-X J0704$-$37, which are conjectured to be on the evolutionary path towards becoming polars (\Fig~\ref{fig:lpt_comparison}), suggesting the possibility that \src{}, too, is such a system.
Notably, its period of ${\sim}1.16\,$hours would make it shorter than the minimum known orbital period of $1.3\,$hours for confirmed polars \citep{schwope2025polarcatcatalogpolarslowaccretion}.
This makes it an interesting test bed for the possibility that LPTs with different periods represent different stages of WD binary evolution, as they progress towards becoming polars.
\begin{figure*}
    \centering
    \includegraphics[width=0.98\linewidth]{lpt_comparison.pdf}
    \caption{Duty cycle vs radio pulse period for various long period radio emitters \citep{deRuiter2025,2022Natur.601..526H,2023Natur.619..487H,2024NatAs...8.1159C,2005Natur.434...50H,deRuiter2025,Lee2025,2024arXiv241116606W}. The duty cycles are derived from the reported pulse widths for LPTs, which are compared to the range of the fitted, unscattered pulse widths of \src{} at 10\% of the peak. AR\,Sco \citep{2016Natur.537..374M,2018A&A...611A..66S} and other sources without well-defined duty cycles or periods \citep[e.g.][and references therein]{2021ApJ...920...45W} are not included on this plot.}
    \label{fig:lpt_comparison}
\end{figure*}

@@

Although there is as yet no clear consensus on whether LPTs represent one class or multiple classes of system \citep[e.g.][]{2024ApJ...961..214R}, the recent confirmation of J1912$-$4410 as a pre-evolved white dwarf pulsar \citep{2023NatAs...7..931P} and ILT\,J1101+5521 and GLEAM-X\,J0704$-$37 as likely candidates for polars or polar-like systems \citep{deRuiter2025,2025A&A...695L...8R} raise the intriguing possibility that other LPTs may also be radio-emitting white dwarf pulsars at various stages of evolution (\citealt{2021NatAs...5..648S}, Horv\'{a}th et al. submitted).

@@

Interestingly, systems with spin-orbital (or beat-orbital) resonances of, e.g., 2:1 or 3:1, would be indistinguishable from systems in 1:1 resonances if pulses can be observed within a limited orbital phase range.
We therefore raise the possibility that the coincidence of the observed radio periodicities of ILT\,J1101+5521 and GLEAM-X\,J0704$-$37 with their spectroscopically confirmed orbital periods does not necessarily imply that they are 1:1 polars.

@@

Another possibility is that radio emission is only observed when the companion produces dense enough particle outflows, in the form of winds, to activate the emission mechanism in the vicinity of the WD.
Such outflows can last for months \citep{2021ApJ...915...37W}, and would not be periodic.
We also note that the intermittency and varying polarisation properties of \src{} (and other LPTs) may be explained by changes in the local plasma environment (Rose et al., in prep).
If intermittency is of this kind, then the observed activity windows will be similarly aperiodic, or quasiperiodic.
Again, long-term monitoring can help distinguish between these scenarios.

@@

The raw visibilities for the MWA, ASKAP, and MeerKAT data sets are available via the respective telescopes' data portals.
The ATCA data is available on reasonable request to the authors, and will become publicly available from the relevant data portal by November, 2025.
The post-imaged dynamic spectra, analyses, and timing analysis data products (ToAs, pulse properties) are publicly available on GitHub\footnote{TODO: Make this available upon publication.}.

@@

>>>>

% Using this for post-review corrections
\newcommand{\revision}[2]{{\color{red}#2}}

\newcommand{\src}{J1755$-$2527}
\newcommand{\srcfull}{ASKAP J175534.9$-$252749.1}

@@

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
%\title[Confirmation of \src{} as LPT]{Confirmation of \srcfull{} as a long period radio transient}
\title[1.16-hr pulses from \src{}]{\revision{Confirmation of \srcfull{} as a long period radio transient}{A new long period radio transient: Discovery of pulses repeating every 1.16 hours from \srcfull{}}}

@@

% Abstract of the paper
\begin{abstract}
We report the discovery of several new pulses from the source \srcfull{} (\src{}), originally identified from a single 2-min long pulse, confirming it as a long period transient (LPT) with a period of ${\sim}1.16\,$hours. The pulses are significantly scattered, consistent with Galactic electron density models. Two of the new pulses also had measurable polarisation, but unlike the originally detected pulse, the polarisation angle does not behave as expected from the rotating vector model. We interpret historical non-detections of \src{} as an intrinsic intermittency that occurs on month-long timescales, and discuss possible causes. We conjecture that, like some other LPTs with periods $\gtrsim 1\,$hour, \src{} \revision{is a white-dwarf binary pulsar, but note that its period is marginally shorter than the minimum known orbital period for the class of cataclysmic variables known as polars}{may host a white dwarf in a binary orbit, but note that its period is marginally shorter than the canonical orbital period minimum of cataclysmic variables}. \revision{}{Our work
highlights the importance of additional observations in establishing the nature of unusual radio-
emitting objects.}
\end{abstract}

@@

\srcfull{} (hereafter, \src{}) is a transient radio source discovered \citep[][hereafter \citetalias{2024MNRAS.535..909D}]{2024MNRAS.535..909D} in the Variables and Slow Transients (VAST) survey \citep{2013PASA...30....6M}.
A single, highly polarised pulse was observed, lasting approximately 2 minutes, exhibiting a steep spectrum ($\alpha = -3.1$, for flux density $S_\nu\propto\nu^\alpha$).
Extensive follow up in new and archival data at multiple wavelengths turned up no other pulses.
Despite this, the authors concluded that \src{} was most likely a long period transient (LPT), a newly emerging class of radio source with periods ranging from minutes to hours \citep[see e.g.][]{2022Natur.601..526H,2023Natur.619..487H,2024NatAs...8.1159C}.
\revision{}{Without additional evidence, however, \src{}
has heretofore not been classified as an LPT in current catalogs.}

Although optical observations have provided compelling evidence that at least some LPTs are detached, possibly strongly magnetic, \revision{white dwarf (WD) binaries}{white dwarf / M-dwarf (WD + MD) binary systems} \citep{deRuiter2025,2024ApJ...976L..21H,2025A&A...695L...8R}, it remains unclear if this model can account for all known LPTs \citep[e.g.][]{2022ApJ...940...72R,Lee2025}.
Optical follow-up is challenging for \src{}, which is situated very close to the Galactic plane ($b = -0\overset{\circ}{.}12$), at a distance of ${\sim}4.7\,$kpc \citepalias{2024MNRAS.535..909D}.
Unsurprisingly, archival searches at \src{}'s location did not turn up any source down to ${\sim}23$ AB mag.
Without more detections in the radio or at other wavelengths, it seemed unlikely that the nature of \src{} could ever be resolved.

@@

Pulses from \src{} were detected with the Murchison Widefield Array \citep[MWA;][]{Tingay2013}, the Australian SKA Pathfinder \citep[ASKAP;][]{2021PASA...38....9H}, MeerKAT \citep{2016mks..confE...1J}, and the Australia Telescope Compact Array \citep[ATCA;][]{cabb}, all radio interferometers capable of imaging.
A summary of all the observations containing pulse detections included in this work is \revision{summarised}{presented} in \Tab~\ref{tbl:obs}.
In this section, we describe how each telescope was used to generate dynamic spectra of \src{}.

@@

\src{} was detected \revision{blindly}{} with the Murchison Widefield Array \citep[MWA;][]{Tingay2013} in observations taken between June--September 2024 under the `Galactic Plane Monitor', project code G0080. The prime objective of the campaign was to search for long-period transients by scanning the Galactic plane over $285^\circ < l < 65^\circ$, $|b| < 15^\circ$ on a bi-weekly cadence at 185--215\,MHz \citep[see Methods of][]{2023Natur.619..487H}; a full description will be released by Hurley-Walker et al., in prep. We also observed over June--September 2022 but made no detections in these data, as reported in \citetalias{2024MNRAS.535..909D}. \revision{}{The root-mean-squared (RMS) noise level in these observations on the processed four-second timescale is $\sim$80\,mJy\,beam$^{-1}$ in this region. A time-based convolution filter employed to extract pulse-like signals, as described by Horv\'ath et al., submitted, yielded the first detection in this data set of a pulse whose peak flux density was 665\,mJy.}
Subsequent to its redetection, we commenced a Director's Discretionary Time (DDT) campaign under project code D0042, to obtain further observations, adjusting the frequency from 200.32\,MHz to 184.96\,MHz for some of the observations to attempt to improve measurements of its DM.
Time of arrival (ToA) predictions for the DDT observations were made using a preliminary ephemeris derived from the GPM detections, which, by virtue at being at the same or similar frequencies, absorbed the delays due to dispersion and scattering into the `PEPOCH' term.

@@

With the re-detection of the source in the MWA data, and the calculation of an ephemeris (\Sect~\ref{sec:analysis}), we re-reduced the data in which we would predict that \src{} would appear. We made one further detection, in SBID~49153, at a relatively low significance ($\sim5\sigma$)\revision{ compared to the blind detections}. We followed a similar procedure as the above to generate dynamic spectra for each of the observations, using a minimum baseline length of 500\,m to suppress uncleaned Galactic diffuse emission.

@@

The correlator data were calibrated using the standard SARAO SDP calibration pipeline, and imaged using \textsc{WSClean}. We formed a deep model of the sources within the primary beam, excepting \src{}, and subtracted this from the visibilities.
We then averaged the visibilities, excluding all baselines shorter than 150\,m, to produce a dynamic spectrum of the source, shown in \Fig~\ref{fig:stacked_spectra}.
\revision{}{We used the source-finding software \texttt{Aegean} \citep{2018PASA...35...11H} to obtain an updated position of 17:55:34.87(2) $-$25:27:49.9(4).
As in \citetalias{2024MNRAS.535..909D}, we find no coincident sources in either optical or near infrared at the same location (see Appendix~\ref{app:localization}).}

@@

These ToAs were then barycentred and fit for both period and DM, resulting in the values given in \Tab~\ref{tbl:ephemeris}.
The residuals are presented in \Fig~\ref{fig:pulse_details} and \Tab~\ref{tbl:toas}, along with other fitted properties of the pulses.
\revision{We also tried fitting a timing model that included a spin-down parameter, and found that it was consistent with zero spin-down to within $1\sigma$, and so is not included in the ephemeris.}{We also tried fitting a timing model that included a spin-down/up parameter (fixing the DM to $733\,{\rm pc}\,{\rm cm}^{-3}$), and found that it was consistent with zero.
However, owing to recent discovery that CHIME/ILT\,J1634+44 is spinning up \citep[$\dot{P} = (-9.03 \pm 0.11) \times 10^{-12}\,{\rm s}\,{\rm s}^{-1}$;][]{2025arXiv250705139D,2025arXiv250705078B}, we report the fitted period derivative value of $(-10 \pm 92) \times 10^{-12}\,{\rm s}\,{\rm s}^{-1}$ instead of only providing an upper limit for the spin-down.}

\begin{table}
  \centering
  \caption{Timing ephemeris for \src{}}
  \label{tbl:ephemeris}
  \begin{tabular}{lc}
    \hline
    Parameter & Value \\
    \hline
    Period (s) & $4186.3285 \pm 0.0002$ \\
    PEPOCH (MJD) & $59965.03792 \pm 0.00003$ \\
    DM (pc cm$^{-3}$) & $733 \pm 22$ \\
    $\tau_{\rm sc,1 GHz}$ (s) & ${\sim}0.07$ \\
    \revision{}{Spindown rate (${\rm s}\,{\rm s}^{-1}$)} & \revision{}{$(-10 \pm 92) \times 10^{-12}$} \\
    \hline
  \end{tabular}
\end{table}

@@

\src{}'s period and duty cycle are not dissimilar to ILT J1101+5521 and GLEAM-X J0704$-$37, \revision{which are conjectured to be on the evolutionary path towards becoming polars}{both WD + MD binary systems} (\Fig~\ref{fig:lpt_comparison}), suggesting the possibility that \src{}, too, is such a system.
Notably, its period of ${\sim}1.16\,$hours would make it shorter than the minimum known orbital period of $1.3\,$hours for \revision{confirmed}{} polars \citep{2009MNRAS.397.2170G,schwope2025polarcatcatalogpolarslowaccretion}.
This makes it an interesting test bed for the possibility that LPTs with different periods represent different stages of \revision{WD binary evolution, as they progress towards becoming polars}{WD + MD binary evolution}.
\begin{figure*}
    \centering
    \includegraphics[width=0.98\linewidth]{lpt_comparison.pdf}
    \caption{Duty cycle vs radio pulse period for various long period radio emitters \citep{deRuiter2025,2022Natur.601..526H,2023Natur.619..487H,2024NatAs...8.1159C,2005Natur.434...50H,deRuiter2025,Lee2025,2024arXiv241116606W}. The duty cycles are derived from the reported pulse widths for LPTs, which are compared to the range of the fitted, unscattered pulse widths of \src{} at 10\% of the peak. AR\,Sco \citep{2016Natur.537..374M,2018A&A...611A..66S} and other sources without well-defined duty cycles or periods \citep[e.g.][and references therein]{2021ApJ...920...45W} are not included on this plot.}
    \label{fig:lpt_comparison}
\end{figure*}

@@

\revision{Although there}{There} is as yet no clear consensus on whether LPTs represent one class or multiple classes of system \citep[e.g.][]{2024ApJ...961..214R}\revision{}{. However}, the recent confirmation of \revision{J1912$-$4410 as a pre-evolved white dwarf pulsar \citep{2023NatAs...7..931P}}{AR\,Scorpii \citep{2016Natur.537..374M}, J1912$-$4410 \citep{2023NatAs...7..931P}, and SDSS\,J230641.47+244055.8 \citep{2025arXiv250620455C} as white dwarf pulsars} and ILT\,J1101+5521 and GLEAM-X\,J0704$-$37 as \revision{likely candidates for polars or polar-like systems}{WD + MD systems} \citep{deRuiter2025,2025A&A...695L...8R}\revision{}{, and CHIME/ILT J1634+44 as a binary hosting a white dwarf \citep[possibly two,][]{2025arXiv250705139D,2025arXiv250705078B}} raise the intriguing possibility that other LPTs may also be radio-emitting white dwarf pulsars at various stages of evolution (\citealt{2021NatAs...5..648S}, Horv\'{a}th et al. submitted).

@@

Interestingly, systems with spin-orbital (or beat-orbital) resonances of, e.g., 2:1 or 3:1, would be indistinguishable from systems in 1:1 resonances if pulses can \revision{}{only} be observed within a limited orbital phase range.
We therefore raise the possibility that the coincidence of the observed radio periodicities of ILT\,J1101+5521 and GLEAM-X\,J0704$-$37 with their spectroscopically confirmed orbital periods does not necessarily imply that they are 1:1 \revision{polars}{systems}.

@@

Another possibility is that radio emission is only observed when the companion produces dense enough particle outflows, in the form of winds, to activate the emission mechanism in the vicinity of the WD.
\revision{}{This may be the case, for example, for GLEAM-X\,J0704$-$37, whose spectroscopically detected H$\alpha$ emission is associated with the M-dwarf, and which may come from its stellar wind \citep{2025A&A...695L...8R}.}
Such outflows can last for months \citep{2021ApJ...915...37W}, and would not be periodic.
We also note that the intermittency and varying polarisation properties of \src{} (and other LPTs) may be explained by changes in the local plasma environment (Rose et al., in prep).
If intermittency is of this kind, then the observed activity windows will be similarly aperiodic, or quasiperiodic.
Again, long-term monitoring can help distinguish between these scenarios.

@@

The raw visibilities for the MWA\footnote{\url{https://asvo.mwatelescope.org/}}, ASKAP\footnote{\url{https://research.csiro.au/casda/}}, and MeerKAT\footnote{\url{https://archive.sarao.ac.za/}} data sets are available via the respective telescopes' data portals.
The ATCA\footnote{\url{https://atoa.atnf.csiro.au/}} data is available on reasonable request to the authors, and will become publicly available from the relevant data portal by November, 2025.
The post-imaged dynamic spectra, analyses, and timing analysis data products (ToAs, pulse properties) are publicly available on GitHub\footnote{TODO: Make this available upon publication.}.

@@

\section{\revision{}{MeerKAT localization}}
\label{app:localization}

\Fig~\ref{fig:localization} shows the position derived from the MeerKAT detection on MJD 60602.

\begin{figure*}
    \centering
    \includegraphics[width=0.48\linewidth]{J1755_optical.pdf}
    \includegraphics[width=0.48\linewidth]{J1755_nir.pdf}
    \caption{The $1\sigma$ positional uncertainty of \src{} derived from the MeerKAT observation taken on MJD 60602 is shown as the green ellipses overlaid on the optical (z, i, r) Panoramic Survey Telescope and Rapid Response System (PanSTARRS) image (left panel), and the infrared (K, H, J) United Kingdom Infrared Telescope Deep Sky Survey (UKIDSS) image (right panel).}
    \label{fig:localization}
\end{figure*}
